---
title: "Demo script for mother machine data analysed with MoMA"
author: Thomas Julou
output: html_document
---


```{r variables, include=FALSE}
dt <- 3             # frame intervall (min)
dl <- 0.065         # pixel size (µm)
vertical_cutoff <- 4 / dl   # after it touched this coordinate a cell is discarded

proj_path <- "~/Documents/Biozentrum/Projects/vngWetLabR/mother_machine/example/"
r_scripts_path <- c("~/Documents/Biozentrum/Projects/vngWetLabR/mother_machine",
                  "~/Documents/Biozentrum/Projects/vngWetLabR/ggplot")
perl_scripts_path <- "~/Documents/Biozentrum/Projects/vngWetLabR/mother_machine"
data2preproc <- function(.d) sub('/data/*', '/preproc/', .d) # store cache file in preproc subdir

condition_ts <- rbind(data.frame(duration=1440, medium='lactose', condition='lac_lowillum') #,
                      # data.frame(duration=c(360, 240, 240, 240, 240, 240),
                      #            medium=c('glucose', 'lactose', 'glucose', 'lactose', 'glucose', 'lactose'),
                      #            condition='switch_4h')
)

date_cond <- c("20160318"="lac_lowillum")

# # save environment (but `pls`)
# myvars <- ls(all.names = TRUE)
# save(list=myvars[which(myvars!='pls')], file=".RData", envir=.GlobalEnv)

```


```{r settings}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE) # , include=FALSE

# set working environment
invisible(sapply(
  list.files(r_scripts_path, pattern="\\.[Rr]$", full.names=TRUE, ignore.case=TRUE), 
  source, .GlobalEnv))
setwd(proj_path)

library(parallel)
numCores <- min(30, detectCores()-1) # do not use more than 30 cores

library(multidplyr)
mycluster <- create_cluster(numCores)
for (.l in mylibs) { mycluster %>% cluster_library(.l) }

```


Load all frames in a dataframe:\
NB: this step calls a perl script to analyze MoMA's output. If all files have already been processed, data will be loaded in a dataframe; otherwise, files to be analyze will be submitted to the cluster's queue. To force the preprocessing to be run again for all files, use .force=TRUE in `process_moma_data`.

```{r}
condition_ts <- condition_ts %>% group_by(condition) %>% 
  mutate(t_start=cumsum(c(0, duration[-(length(duration))])) * 60,
         t_end=cumsum(duration) * 60 - 1e-5,
         m_cycle=value_occurence_index(medium))

myfiles <- list.files("./data", ".*\\d+\\.csv", recursive=TRUE, full.names=TRUE)
nc <- min(length(myfiles), length(mycluster)) # this is a dirty hack because multiplyr crashes with less shards than cores

# load perl scripts output to dataframes (using parallel lapply)
myframes <- myfiles %>% 
  # process exported files on the cluster if required
  process_moma_data(.data2preproc=data2preproc, .scripts_path=perl_scripts_path, 
                    .qsub_name="MMex_pl", .force=FALSE) %>% 
  dplyr::data_frame(path=.) %>% 
  mutate(ppath=data2preproc(path)) %>% 
  # load perl scripts output to dataframes (using multidplyr)
  # group_by(path) %>%
  partition(path, cluster=mycluster[1:nc] %>%
              cluster_assign_func(load_moma_processed, parse_frames_stats, compute_genealogy, which_touch_exit, which_to_progeny) %>%
              cluster_assign_obj(dt, dl, vertical_cutoff)) %>%
  do((function(.df){
    load_moma_processed(.df$ppath) %>% 
      mutate(time_sec=frame*dt*60, length_um=length_pixel*dl,
             discard_start=(time_sec < 2*3600)) %>%
      # remove frames after touching the exit
      group_by(id) %>%
      mutate(discard_top=which_touch_exit(vertical_top, vertical_cutoff)) %>%
      mutate(discard_top=ifelse(discard_start, FALSE, discard_top)) %>% # not in the preexpt step (2h)
      mutate(end_type=ifelse(any(discard_top), 'exit', end_type)) %>% # update end_type
      # remove daughters of cells that touched the exit
      ungroup %>%
      mutate(discard_top=which_to_progeny(discard_top, cid))
  })(.)) %>%
  collect()
  group_by(date, pos, gl, id) %>%
  mutate(start_time=first(time_sec), end_time=last(time_sec),
         b_rank=round(mean(total_cell_in_lane - cell_num_in_lane))) %>% 
  #   filter(discard_top==FALSE) %>%
  ungroup %>% 
  mutate(condition=date_cond[as.character(date)],
         vertical_center=(vertical_bottom + vertical_top)/2,
         cell=paste(date, pos, gl, id, sep='.')) %>% 
  # propagate medium info
  group_by(condition) %>% 
  do((function(.df){
    .ts <- filter(condition_ts, condition==unique(.df$condition))
    .idx <- find_unique_interval(.df$time_sec, .ts$t_start, .ts$t_end)
    mutate(.df, medium=.ts$medium[.idx], m_start=.ts$t_start[.idx], m_cycle=.ts$m_cycle[.idx])
  })(.))  

autofluo_per_um <- 422.8 / 8
fp_per_dn <- 1

autofluo_predict <- function(.h) .h * autofluo_per_um
myframes <- myframes %>%
  mutate(fluogfp_amplitude = fluo_amplitude - autofluo_predict(length_um),
         gfp_nb = fluogfp_amplitude * fp_per_dn)

```


Compute statistics per cell:

```{r}
mycells <- filter(myframes, condition!='switch', !discard_top) %>%
  # group_by(date, pos, gl, id, parent_id, genealogy) %>%
  partition(condition, date, pos, gl, id, parent_id, genealogy, cluster=mycluster) %>%
  filter(!any(discard_start), end_type=='div') %>% # full cell cycles only
  filter(n()>4) %>% # at least 4 time points
  do((function(.df) {
    # browser()
    
    .mod_ll_t <- lm( log(length_um)~time_sec, .df)  # use fastLm() for predict
    # .mod_ll_t <- fastLmPure( cbind(1, .df$time_sec), log(.df$length_um) )
    .mod_lg_t <- fastLmPure( cbind(1, .df$time_sec), log(.df$gfp_nb) )
    .mod_l_t <- fastLmPure( cbind(1, .df$time_sec), .df$length_um )
    .mod_g_t <- fastLmPure( cbind(1, .df$time_sec), .df$gfp_nb )
    .mod_g_l <- fastLmPure( cbind(1, .df$length_um), .df$gfp_nb )

    .time_birth <- first(.df$time_sec)
    .time_div <- last(.df$time_sec)
    .logl <- predict(.mod_ll_t, se.fit=TRUE)
    data.frame(npoints=.mod_ll_t$df.residual+1,
               time_birth=.time_birth, time_div=.time_div, 
               cell_num_from_top=mean(.df$cell_num_in_lane),
               cell_num_from_bottom=mean(.df$total_cell_in_lane-.df$cell_num_in_lane), 
               loglength_start=first(.logl$fit), loglength_startse=first(.logl$se.fit), 
               loglength_end=last(.logl$fit), loglength_endse=last(.logl$se.fit), 
               logl_time_slope=.mod_ll_t$coefficients[2], logl_time_slopesd=summary(.mod_ll_t)$coefficients[2,2], 
               # logl_time_slope=.mod_ll_t$coefficients[2], logl_time_slopesd=.mod_ll_t$stderr[2], 
               logl_time_r2=corPearson(.df$time_sec, log(.df$length_um))^2,
               logg_time_slope=.mod_lg_t$coefficients[2], logg_time_slopesd=.mod_lg_t$stderr[2], 
               logg_time_r2=corPearson(.df$time_sec, log(.df$gfp_nb))^2,
               l_time_slope=.mod_l_t$coefficients[2], l_time_slopesd=.mod_l_t$stderr[2], 
               l_time_r2=corPearson(.df$time_sec, .df$length_um)^2,
               g_time_slope=.mod_g_t$coefficients[2], g_time_slopesd=.mod_g_t$stderr[2], 
               g_time_r2=corPearson(.df$time_sec, .df$gfp_nb)^2,
               g_l_slope=.mod_g_l$coefficients[2], g_l_slopesd=.mod_g_l$stderr[2], 
               g_l_r2=corPearson(.df$length_um, log(.df$gfp_nb))^2)
  })(.) ) %>% 
  collect() %>% 
  arrange(condition, date, pos, gl, id)

print(mycells)
```


Plot cell traces for length and fluo.

Minimal example for one growth lane:

```{r}
myframes %>% 
  filter((date=='20160318' & pos==0 & gl==2)) %>% 
  filter(!discard_top, vertical_top>vertical_cutoff) %>% 
  mutate(time_sec=time_sec-2*3600) %>% 
  plot_faceted_var_tracks(.var_col='length_um', .show_cellid=TRUE, .log=TRUE) +
  # mask early frames (requires a dummy df!)
  geom_rect(aes(xmin=-Inf, xmax=0, ymin=0, ymax=Inf, group=1), fill='white', alpha=.6, col='transparent', data=data.frame(a=1)) +
  scale_x_hours(4) +
  scale_y_continuous(breaks=2:4, trans='log2') +
  labs(y='length (µm)')

```

for all growth lanes, several variables, saved as multipage pdf:

```{r eval=FALSE}
pls <- myframes %>% 
  group_by(condition, date, pos, gl) %>%
  do(pll=(function(.df){
    # browser()
    if (dim(filter(.df, !discard_top))[1] == 0) return(list())
    .cond <- unique(.df$condition)
    .fill <- RColorBrewer::brewer.pal(3, 'Set1')
    .hmin <- max(1.2,  min(filter(.df, !discard_top)$length_um) )
    .hrange <- (max(filter(.df, !discard_top)$length_um)-.hmin) / 2
    .fmin <- min(filter(.df, !discard_top)$gfp_nb)
    .frange <- (max(filter(.df, !discard_top)$gfp_nb)-.fmin) / 2
    
    custom_labels <- function (.str) {
      .labels <- paste('rank:', .str)
      .labels[.labels=='rank: -1'] <- 'all'
      .labels[.labels=='rank: 6'] <- 'rank: >= 6'
      return(.labels)
    }
    
    list(
      l = filter(.df, !discard_top, vertical_top>vertical_cutoff) %>% 
        mutate(time_sec=time_sec-2*3600) %>% 
        plot_faceted_var_tracks(.var_col='length_um', .show_all=TRUE, .show_cellid=TRUE, .log=TRUE, .facet_labeller=custom_labels) +
        # show medium bar
        geom_rect(aes(xmin=t_start - 2*3600, xmax=t_end - 2*3600, ymin=hmin, ymax=hmax, fill=medium, group=1), size=0.2, data=filter(condition_ts, condition==.cond) %>% mutate(b_rank=-1, hmin=.hmin-.25, hmax=.hmin-.1)) +
        geom_vline(aes(xintercept=t_start - 2*3600, group=1), alpha=.2, size=.5, data=filter(condition_ts, condition==.cond)) +
        geom_text(aes(x=t_start+(t_end-t_start)/2 - 2*3600, y=h, label=medium, group=1), col='white', size=2, hjust=0.5, vjust=0,
                  data=filter(condition_ts, condition==.cond) %>% mutate(b_rank=-1, h=.hmin-.2)) +
        scale_fill_manual(values=c('glucose'=.fill[1], 'lactose'=.fill[2]), guide='none') +
        # mask early frames (requires a dummy df!)
        geom_rect(aes(xmin=-Inf, xmax=0, ymin=0, ymax=Inf, group=1), fill='white', alpha=.6, col='transparent', data=data.frame(a=1)) +
        scale_x_hours(4) +
        scale_y_continuous(breaks=2:4, trans='log2') +
        labs(y='length (µm)'), 
      
      ft = filter(.df, !discard_top, vertical_top>vertical_cutoff) %>% 
        mutate(time_sec=time_sec-2*3600) %>% 
        plot_faceted_var_tracks(.var_col='gfp_nb', .show_all=TRUE, .show_cellid=TRUE, .facet_labeller=custom_labels) +
        # show medium bar
        geom_rect(aes(xmin=t_start - 2*3600, xmax=t_end - 2*3600, ymin=fmin, ymax=fmax, fill=medium), size=0.2, data=filter(condition_ts, condition==.cond) %>% mutate(b_rank=-1, fmin=.fmin-.frange/2.5, fmax=.fmin-.frange/5)) +
        geom_vline(aes(xintercept=t_start - 2*3600), alpha=.2, size=.5, data=filter(condition_ts, condition==.cond, t_start>0)) +
        geom_text(aes(x=(t_start+(t_end-t_start)/2 - 2*3600), y=f, label=medium), col='white', size=2, hjust=0.5, vjust=0,
                  data=filter(condition_ts, condition==.cond) %>% mutate(b_rank=-1, f=.fmin-.frange/3)) +
        scale_fill_manual(values=c('glucose'=.fill[1], 'lactose'=.fill[2]), guide='none') +
        # mask early frames (requires a dummy df!)
        geom_rect(aes(xmin=-Inf, xmax=0, ymin=-Inf, ymax=Inf, group=1), fill='white', alpha=.6, col='transparent', data=data.frame(a=1)) +
        scale_x_hours(4) +
        # scale_y_continuous(breaks=2:4, trans='log2') +
        labs(y='total GFP per cell (molecules)'),
      
      fc = filter(.df, !discard_top, vertical_top>vertical_cutoff) %>% 
        mutate(time_sec=time_sec-2*3600, gfp_conc=gfp_nb/length_um) %>% 
        plot_faceted_var_tracks(.var_col='gfp_conc', .show_all=TRUE, .show_cellid=TRUE, .facet_labeller=custom_labels) +
        # show medium bar
        geom_rect(aes(xmin=t_start - 2*3600, xmax=t_end - 2*3600, ymin=fmin/2, ymax=fmax/2, fill=medium), size=0.2, data=filter(condition_ts, condition==.cond) %>% mutate(b_rank=-1, fmin=.fmin-.frange/2.5, fmax=.fmin-.frange/5)) +
        geom_vline(aes(xintercept=t_start - 2*3600), alpha=.2, size=.5, data=filter(condition_ts, condition==.cond, t_start>0)) +
        geom_text(aes(x=(t_start+(t_end-t_start)/2-120), y=f/2, label=medium), col='white', size=2, hjust=0.5, vjust=0, 
                  data=filter(condition_ts, condition==.cond) %>% mutate(b_rank=-1, f=.fmin-.frange/3)) +
        scale_fill_manual(values=c('glucose'=.fill[1], 'lactose'=.fill[2]), guide='none') +
        # mask early frames (requires a dummy df!)
        geom_rect(aes(xmin=-Inf, xmax=0, ymin=-Inf, ymax=Inf, group=1), fill='white', alpha=.6, col='transparent', data=data.frame(a=1)) +
        scale_x_hours(4) +
        # scale_y_continuous(breaks=2:4, trans='log2') +
        labs(y='total GFP per cell (molecules)') 
    )
  })(.))

pdf('plots/example_path_length.pdf', width=12, height=10)
for (i in 1:dim(pls)[1]) {
  if ('l' %in% names(pls[[i, 'pll']]))
    plot(pls[[i, 'pll']] [['l']] + 
           labs(title=sprintf("%s  pos:%02d  GL:%02d", pls[[i, "date"]], pls[[i, "pos"]], pls[[i, "gl"]])))
}
dev.off()

pdf('plots/example_path_fluotot.pdf', width=12, height=10)
for (i in 1:dim(pls)[1]) {
  if ('ft' %in% names(pls[[i, 'pll']]))
  plot(pls[[i, 'pll']] [['ft']] + 
         labs(title=sprintf("%s  pos:%02d  GL:%02d", pls[[i, "date"]], pls[[i, "pos"]], pls[[i, "gl"]])))
}
dev.off()

pdf('plots/example_path_fluoconc.pdf', width=12, height=10)
for (i in 1:dim(pls)[1]) {
  if ('fc' %in% names(pls[[i, 'pll']]))
  plot(pls[[i, 'pll']] [['fc']] + 
         labs(title=sprintf("%s  pos:%02d  GL:%02d", pls[[i, "date"]], pls[[i, "pos"]], pls[[i, "gl"]])))
}
dev.off()

```

